---
title: "Data Preperation"
author: "Siyang Ni"
date: "2024-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Description

The dataset is from a longgitudinal social survey in the UK called the "Millennium Cohort Study" (MCS). The MCS follows the lives of around 19,000 young people born in England, Scotland, Wales, and Northern Ireland in 2000-02. 

MCS provides multiple measures of the cohort members’ physical, socio-emotional, cognitive and behavioral development over time, as well as detailed information on their daily life, behavior and experiences. Alongside this, rich information on economic circumstances, parenting, relationships and family life is available from both resident parents.

Currently, the MCS have seven waves available to researchers. The first wave of survey was taken at the year 2001, when respondents were nine-month old. The seventh wave survey was taken at 2018, when the respondents were seventeen-year-old.   


## Tasks

We will conduct the following task:

1. Merge the seven waves of data.
2. Provide a general description of the contents of your sample.
3. Identify issues that may need to be addressed in later analyses.
4. Exploratory data visualization 
5. Some very basic exploratory data analysis

Computing Environment: 

-Program: R 4.3.2
-IDE: R Studio Server 2023.12.1+402 "Ocean Storm" Release 
-OS: Ubuntu 22.04.4 LTS through WSL2 (x86_64-pc-linux-gnu)


## Loading and Merging 

```{r}
setwd("/home/freeyang/research/hdfs597")

# install.packages('pacman')
# Pacman is an efficient R package manager, works like APT in Debian
library(pacman) 

# For merging we need haven and dplyrgit config --global user.email "youremail@domain.com"
pacman::p_load(haven, dplyr)
```

```{r, eval=FALSE}
# Load all waves

## Base directory prefix
base_dir <- "/home/freeyang/research_data/mcs"

# Reading each wave, all column names to lowercase, filter out twins and triplets, keep only main caregivers. 

# Wave 1 Parent Interview
wave1_parent_interview <- read_dta(file.path(base_dir, "MCS 1/stata11/mcs1_parent_interview.dta")) %>%
  rename_with(tolower)

# Wave 2 Child Assessment
wave2_child_assessment <- read_dta(file.path(base_dir, "MCS 2/stata11_se/mcs2_child_assessment_data.dta")) %>%
  rename_with(tolower) %>%
  filter(bhcnum00 == 1)

# Wave 2 Parent Interview
wave2_parent_interview <- read_dta(file.path(base_dir, "MCS 2/stata11_se/mcs2_parent_interview.dta")) %>%
  rename_with(tolower)

# Wave 3 Child Assessment
wave3_child_assessment <- read_dta(file.path(base_dir, "MCS 3/stata11_se/mcs3_child_assessment_data.dta")) %>%
  rename_with(tolower) %>%
  filter(chcnum00 == 1)

# Wave 3 Parent Interview
wave3_parent_interview <- read_dta(file.path(base_dir, "MCS 3/stata11_se/mcs3_parent_interview.dta")) %>%
  rename_with(tolower)

# Wave 4 Parent Interview
wave4_parent_interview <- read_dta(file.path(base_dir, "MCS 4/stata11_se/mcs4_parent_interview.dta")) %>%
  rename_with(tolower)

# Wave 4 Child Assessment
wave4_child_assessment <- read_dta(file.path(base_dir, "MCS 4/stata11_se/mcs4_cm_assessment.dta")) %>%
  rename_with(tolower) %>%
  filter(dccnum00 == 1)

# Wave 5 CM Assessment
wave5_cm_assessment <- read_dta(file.path(base_dir, "MCS 5/stata11/mcs5_cm_cognitive_assessment.dta")) %>%
  rename_with(tolower) %>%
  filter(ecnum00 == 1)

# Wave 5 CM Interview
wave5_cm_interview <- read_dta(file.path(base_dir, "MCS 5/stata11/mcs5_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(ecnum00 == 1)

# Wave 5 Parent CM Interview
wave5_parent_cm_interview <- read_dta(file.path(base_dir, "MCS 5/stata11/mcs5_parent_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(ecnum00 == 1, epnum00 == 1)

# Wave 5 Child Derived
wave5_child_derived <- read_dta(file.path(base_dir, "MCS 5/stata11/mcs5_cm_derived.dta")) %>%
  rename_with(tolower) %>%
  filter(ecnum00 == 1)

# Wave 6 CM Interview
wave6_cm_interview <- read_dta(file.path(base_dir, "MCS 6/stata11/mcs6_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(fcnum00 == 1)

# Wave 6 Parent CM Interview (Main)
wave6_parent_cm_interview <- read_dta(file.path(base_dir, "MCS 6/stata11/mcs6_parent_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(fpnum00 == 1, fcnum00 == 1)

# Wave 6 Parent Derived
wave6_parent_derived <- read_dta(file.path(base_dir, "MCS 6/stata11/mcs6_parent_derived.dta")) %>%
  rename_with(tolower) %>%
  filter(fpnum00 == 1)

# Wave 6 Child Derived
wave6_child_derived <- read_dta(file.path(base_dir, "MCS 6/stata11/mcs6_cm_derived.dta")) %>%
  rename_with(tolower) %>%
  filter(fcnum00 == 1)

# Wave 7 CM Interview
wave7_cm_interview <- read_dta(file.path(base_dir, "MCS 7/stata13/mcs7_cm_interview_onlyCM1.dta")) %>%
  rename_with(tolower)

# Wave 7 Parent CM Interview
wave7_parent_cm_interview <- read_dta(file.path(base_dir, "MCS 7/stata13/mcs7_parent_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(gpnum00 == 1, gcnum00 == 1)

# Wave 7 CM Derived
wave7_cm_derived <- read_dta(file.path(base_dir, "MCS 7/stata13/mcs7_cm_derived_onlyCM1.dta")) %>%
  rename_with(tolower)
```

```{r, eval=FALSE}
# Merge all waves

## purrr package for merging
pacman::p_load(purrr)

## Store all waves in a list
data_frames_list <- list(
  wave1_parent_interview,
  wave2_child_assessment,
  wave2_parent_interview,
  wave3_child_assessment,
  wave3_parent_interview,
  wave4_parent_interview,
  wave4_child_assessment,
  wave5_cm_assessment,
  wave5_cm_interview,
  wave5_parent_cm_interview,
  wave5_child_derived,
  wave6_cm_interview,
  wave6_parent_cm_interview,
  wave6_parent_derived,
  wave6_child_derived,
  wave7_cm_interview,
  wave7_parent_cm_interview,
  wave7_cm_derived
)

## Use reduce from purrr to merge all waves 1 by 1 
merged_data <- purrr::reduce(data_frames_list, full_join, by = "mcsid")

## Save the merged dataset as an RDS file
### (Uncomment when needed)
#saveRDS(merged_data, file = "/home/freeyang/research_data/mcs/merging/merged_dataset.rds")
```


## Data Exploration

A quick look shows we have a very large dataset with 19,243 rows and 18,816 columns, and also many missing cases. This is expected for a panel data that span from when the respondents were nine-month-old to seventeen-year-old.

```{r}
## Load the merged dateset
merged_data <- readRDS(file = "/home/freeyang/research_data/mcs/merging/merged_dataset.rds")

## Summarize the data
dim(merged_data)
head(merged_data[1:5, 1:10])

## Missing Values
total_missing_values <- sum(is.na(merged_data))
cat("Total Missing Values:", total_missing_values, "\n")
```


Our main target variable is adolescent aggressive behaviors. We look at these variables in each waves, strating with the most recent wave 7 (age 17):

```{r}
# Show frequency distribution for variables measuring aggression at age 17
aggg_7 <- c("gpsdfb00", "gchitt00", "gcwepn00", "gcsdql00")

freq_tables <- lapply(merged_data[aggg_7], table)

lapply(names(freq_tables), function(x) {
  cat("\nFrequency table for", x, ":\n")
  print(freq_tables[[x]])
})
```

In the following code chunk, we recode these variables by comining missing value categories and adding variable labels. Finally, we show the frequency of each variables again. 

```{r}
# Package for label and the tab1 function
pacman::p_load(labelled, epiDisplay)

# Function to apply the recoding based on specific rules
recode_var <- function(var, rules) {
  case_when(
    var %in% rules$from ~ rules$to[match(var, rules$from)],
    TRUE ~ NA_real_  # Default case to NA for unspecified values
  )
}

# Recoding for each variables
df1 <- merged_data %>%
  dplyr::select(gpsdfb00, gchitt00, gcwepn00, gcsdql00) %>% 
  mutate(
    agr7_fig = recode_var(gpsdfb00, tibble(from = c(-1, 4, 1, 2, 3), to = c(NA, NA, 0, 1, 2))),
    agr7_hit = recode_var(gchitt00, tibble(from = c(3, 4, 5, 2, 1), to = c(NA, NA, NA, 0, 1))),
    agr7_wepn = recode_var(gcwepn00, tibble(from = c(3, 4, 5, 2, 1), to = c(NA, NA, NA, 0, 1))),
    agr7_srfig = recode_var(gcsdql00, tibble(from = c(-9, -1, 1, 2, 3), to = c(NA, NA, 0, 1, 2)))
  ) %>% 
  dplyr::select(agr7_fig, agr7_hit, agr7_wepn, agr7_srfig) %>% 
  labelled::set_variable_labels(
    agr7_fig = "Fight with others",
    agr7_hit = "Pushed or shoved/hit/slapped/punched someone",
    agr7_wepn = "Used or hit someone with a weapon",
    agr7_srfig = "I fight a lot"
  )
  
# Variables to analyze
agr7 <- c("agr7_fig", "agr7_hit", "agr7_wepn", "agr7_srfig")

# A quick look at each variable
lapply(agr7, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  tab1(df1[, var_name], main = var_name)
})
```


Now, we look at variables for adolescent aggresive behaviors in wave 6 (age 14).

```{r}
# Show frequency distribution for variables measuring aggression at age 14
aggg_6 <- c("fpsdfb00", "fchitt00", "fcwepn00", "fcbulp00", "fcpckp00")

freq_tables <- lapply(merged_data[aggg_6], table)

lapply(names(freq_tables), function(x) {
  cat("\nFrequency table for", x, ":\n")
  print(freq_tables[[x]])
})
```


```{r}
# Recoding for each variables
df2 <- merged_data %>% 
  mutate(
    agr6_fig = recode_var(fpsdfb00, tibble(from = c(-1, -9, 1, 2, 3), to = c(NA, NA, 0, 1, 2))),
    agr6_hit = recode_var(fchitt00, tibble(from = c(-9, -8, -1, 2, 1), to = c(NA, NA, NA, 0, 1))),
    agr6_wepn = recode_var(fcwepn00, tibble(from = c(-9, -8, -1, 2, 1), to = c(NA, NA, NA, 0, 1))),
    agr6_hps = recode_var(fcbulp00, tibble(from = c(-9, -8, -1, 7, 6, 5, 4, 3, 2, 1), to = c(NA, NA, NA, NA, 0, 1, 2, 3, 4, 5))),
    agr6_hpc = recode_var(fcpckp00, tibble(from = c(-9, -8, -1, 6, 5, 4, 3, 2, 1), to = c(NA, NA, NA, 0, 1, 2, 3, 4, 5))),
  ) %>% 
  labelled::set_variable_labels(
    agr6_fig = "Fight with others",
    agr6_hit = "Pushed or shoved/hit/slapped/punched someone",
    agr6_wepn = "Used or hit someone with a weapon",
    agr6_hps = "Hurt or pick on siblings",
    agr6_hpc = "Hurt or pick on other children") %>%
  dplyr::select(agr6_fig, agr6_hit, agr6_wepn, agr6_hps, agr6_hpc)
  
# Variables to analyze
agr6 <- c("agr6_fig", "agr6_hit", "agr6_wepn", "agr6_hps", "agr6_hpc")

# A quick look at each variable
lapply(agr6, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  tab1(df2[, var_name], main = var_name)
})
```


Following the same logic, we take a look at the variables in wave 5 (age 11)

```{r}
# Aggression variables at age 14
aggg_5 <- c("epsdfb00", "ecq55x00", "ecq57x00")

freq_tables <- lapply(merged_data[aggg_5], table)

# Print frequency tables for each aggression variable
lapply(names(freq_tables), function(x) {
  cat("\nFrequency table for", x, ":\n")
  print(freq_tables[[x]])
})
```

And then recode the three variables measuring adolescent aggression at wave 5 (age 11).

```{r}
df3 <- merged_data %>%
  dplyr::select(epsdfb00, ecq55x00, ecq57x00) %>%
  mutate(
    # Recoding 'fight with others'
    agr5_fig = recode_var(epsdfb00, tibble(from = c(-1, 4, 1, 2, 3), to = c(NA, NA, 0, 1, 2))),
    # Recoding 'hurt or pick on siblings'
    agr5_hps = recode_var(ecq55x00, tibble(from = c(-8, 7, 6, 5, 4, 3, 2, 1), to = c(NA, NA, 0, 1, 2, 3, 4, 5))),
    # Recoding 'hurt or pick on other children'
    agr5_hpc = recode_var(ecq57x00, tibble(from = c(-8, 6, 5, 4, 3, 2, 1), to = c(NA, 0, 1, 2, 3, 4, 5)))
  ) %>%
  # Selecting the newly recoded variables for further analysis or inspection
  dplyr::select(agr5_fig, agr5_hps, agr5_hpc) %>% 
  set_variable_labels(
  agr5_fig = "Fight with others",
  agr5_hps = "Hurt or pick on siblings",
  agr5_hpc = "Hurt or pick on other children"
)

# Show all variable frequency 
agr5 <- c("agr5_fig", "agr5_hps", "agr5_hpc")

lapply(agr5, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  tab1(df3[, var_name], main = var_name)
})

```

Now, combine all the adolescent aggression variables into one dataframe:

```{r}
df_aggr <- bind_rows(df1, df2, df3)
# Check dimensions
dim(df_aggr)
```


We ask how self-control affects adolescent aggressive behaviors. So, we need to look at self-control variables. In our study, we measure self-control from two dimensions: impulsivivity and unhealthy risk-taking behaviors. We look at hese variables in each wave. 

We start by looking at wave 7 (age 17)

```{r}
df4 <- merged_data %>%
  mutate(
    imp7_tem = case_when(
      gpsdtt00 %in% c(-1, 4) ~ NA_real_,
      gpsdtt00 == 1 ~ 0,
      gpsdtt00 == 2 ~ 1,
      gpsdtt00 == 3 ~ 2
    ),
    imp7_res = case_when(
      gpsdro00 %in% c(-1, 4) ~ NA_real_,
      gpsdro00 == 1 ~ 0,
      gpsdro00 == 2 ~ 1,
      gpsdro00 == 3 ~ 2
    ),
    imp7_fid = case_when(
      gpsdfs00 %in% c(-1, 4) ~ NA_real_,
      gpsdfs00 == 1 ~ 0,
      gpsdfs00 == 2 ~ 1,
      gpsdfs00 == 3 ~ 2
    ),
    imp7_dis = case_when(
      gpsddc00 %in% c(-1, 4) ~ NA_real_,
      gpsddc00 == 1 ~ 0,
      gpsddc00 == 2 ~ 1,
      gpsddc00 == 3 ~ 2
    ),
    imp7_thi = case_when(
      gpsdst00 %in% c(-1, 4) ~ NA_real_,
      gpsdst00 == 1 ~ 2,
      gpsdst00 == 2 ~ 1,
      gpsdst00 == 3 ~ 0
    ),
    imp7_tas = case_when(
      gpsdte00 %in% c(-1, 4) ~ NA_real_,
      gpsdte00 == 1 ~ 2,
      gpsdte00 == 2 ~ 1,
      gpsdte00 == 3 ~ 0
    ),
    risk7_lie = case_when(
      gpsdoa00 %in% c(-1, 4) ~ NA_real_,
      gpsdoa00 == 1 ~ 0,
      gpsdoa00 == 2 ~ 1,
      gpsdoa00 == 3 ~ 2
    ),
    risk7_ste = case_when(
      gpsdcs00 %in% c(-1, 4) ~ NA_real_,
      gpsdcs00 == 1 ~ 0,
      gpsdcs00 == 2 ~ 1,
      gpsdcs00 == 3 ~ 2
    ),
    risk7_obe = case_when(
      gpsdor00 %in% c(-1, 4) ~ NA_real_,
      gpsdor00 == 1 ~ 2,
      gpsdor00 == 2 ~ 1,
      gpsdor00 == 3 ~ 0
    ),
    risk7_shl = case_when(
      gcstol00 %in% c(3, 4, 5) ~ NA_real_,
      gcstol00 == 2 ~ 0,
      gcstol00 == 1 ~ 1
    ),
    risk7_wpp = case_when(
      gcspry00 %in% c(3, 4, 5) ~ NA_real_,
      gcspry00 == 2 ~ 0,
      gcspry00 == 1 ~ 1
    ),
    risk7_smk = case_when(
      gcsmok00 %in% c(7, 8, 9) ~ NA_real_,
      gcsmok00 == 1 ~ 0,
      gcsmok00 %in% c(2, 3, 4, 5, 6) ~ 1
    ),
    risk7_drk = case_when(
      gcalcd00 %in% c(3, 4, 5) ~ NA_real_,
      gcalcd00 == 2 ~ 0,
      gcalcd00 == 1 ~ 1
    ),
    risk7_damg = case_when(
      gcdamg00 %in% c(3, 4, 5) ~ NA_real_,
      gcdamg00 == 2 ~ 0,
      gcdamg00 == 1 ~ 1
    )
  ) %>% 
  dplyr::select(imp7_tem, imp7_res, imp7_fid, imp7_dis, imp7_thi, imp7_tas, risk7_lie, risk7_ste, risk7_obe, risk7_shl, risk7_wpp, risk7_smk, risk7_drk, risk7_damg) %>% 
  labelled::set_variable_labels(
    imp7_tem = "Tempur Tantrums",
    imp7_res = "Restless",
    imp7_fid = "Fidgeting",
    imp7_dis = "Easily Distracted",
    imp7_thi = "Think Before Act (Reverse Scale)",
    imp7_tas = "Sees Task Through to the End (Reverse Scale)",
    risk7_lie = "Lie or Cheat",
    risk7_ste = "Steal",
    risk7_obe = "Generally Obedient (Reverse Scale)",
    risk7_shl = "Shoplifting",
    risk7_wpp = "Write/Paint in Public Places",
    risk7_smk = "Underage Smoking",
    risk7_drk = "Underage Drinking",
    risk7_damg = "Damage Something That Doesn't Belong to Them"
  )


# Show descriptive statistics 

self_control7 <- c("imp7_tem", "imp7_res", "imp7_fid", "imp7_dis", "imp7_thi", "imp7_tas", "risk7_lie", "risk7_ste", "risk7_obe", "risk7_shl", "risk7_wpp", "risk7_smk", "risk7_drk", "risk7_damg")

lapply(self_control7, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  tab1(df4[, var_name], main = var_name)
})

```


We do the same for wave 6 (age 14):

```{r}
df5 <- merged_data %>%
  mutate(
    # Self-Control Variables
    imp6_tem = case_when(
      fpsdtt00 %in% c(-9, -1) ~ NA_real_,
      fpsdtt00 == 1 ~ 0,
      fpsdtt00 == 2 ~ 1,
      fpsdtt00 == 3 ~ 2
    ),
    imp6_res = case_when(
      fpsdro00 %in% c(-9, -1) ~ NA_real_,
      fpsdro00 == 1 ~ 0,
      fpsdro00 == 2 ~ 1,
      fpsdro00 == 3 ~ 2
    ),
    imp6_fid = case_when(
      fpsdfs00 %in% c(-9, -1) ~ NA_real_,
      fpsdfs00 == 1 ~ 0,
      fpsdfs00 == 2 ~ 1,
      fpsdfs00 == 3 ~ 2
    ),
    imp6_dis = case_when(
      fpsddc00 %in% c(-9, -1) ~ NA_real_,
      fpsddc00 == 1 ~ 0,
      fpsddc00 == 2 ~ 1,
      fpsddc00 == 3 ~ 2
    ),
    imp6_thi = case_when(
      fpsdst00 %in% c(-9, -1) ~ NA_real_,
      fpsdst00 == 1 ~ 2,
      fpsdst00 == 2 ~ 1,
      fpsdst00 == 3 ~ 0
    ),
    imp6_tas = case_when(
      fpsdte00 %in% c(-9, -1) ~ NA_real_,
      fpsdte00 == 1 ~ 2,
      fpsdte00 == 2 ~ 1,
      fpsdte00 == 3 ~ 0
    ),
    # Risk Taking Variables
    risk6_lie = case_when(
      fpsdoa00 %in% c(-9, -1) ~ NA_real_,
      fpsdoa00 == 1 ~ 0,
      fpsdoa00 == 2 ~ 1,
      fpsdoa00 == 3 ~ 2
    ),
    risk6_ste = case_when(
      fpsdcs00 %in% c(-9, -1) ~ NA_real_,
      fpsdcs00 == 1 ~ 0,
      fpsdcs00 == 2 ~ 1,
      fpsdcs00 == 3 ~ 2
    ),
    risk6_obe = case_when(
      fpsdor00 %in% c(-9, -1) ~ NA_real_,
      fpsdor00 == 1 ~ 2,
      fpsdor00 == 2 ~ 1,
      fpsdor00 == 3 ~ 0
    ),
    risk6_shl = case_when(
      fcstol00 %in% c(-9, -8, -1) ~ NA_real_,
      fcstol00 == 2 ~ 0,
      fcstol00 == 1 ~ 1
    ),
    risk6_wpp = case_when(
      fcspry00 %in% c(-9, -8, -1) ~ NA_real_,
      fcspry00 == 2 ~ 0,
      fcspry00 == 1 ~ 1
    ),
    risk6_smk = case_when(
      fcsmok00 %in% c(-9, -8, -1) ~ NA_real_,
      fcsmok00 == 1 ~ 0,
      fcsmok00 %in% c(2, 3, 4, 5, 6) ~ 1
    ),
    risk6_drk = case_when(
      fcalcd00 %in% c(-9, -8, -1) ~ NA_real_,
      fcalcd00 == 2 ~ 0,
      fcalcd00 == 1 ~ 1
    ),
    risk6_rude = case_when(
      fcrude00 %in% c(-9, -8, -1) ~ NA_real_,
      fcrude00 == 2 ~ 0,
      fcrude00 == 1 ~ 1
    ),
    risk6_damg = case_when(
      fcdamg00 %in% c(-9, -8, -1) ~ NA_real_,
      fcdamg00 == 2 ~ 0,
      fcdamg00 == 1 ~ 1
    )
  ) %>%
   dplyr::select(imp6_tem, imp6_res, imp6_fid, imp6_dis, imp6_thi, imp6_tas, risk6_lie, risk6_ste, risk6_obe, risk6_shl, risk6_wpp, risk6_smk, risk6_drk, risk6_rude, risk6_damg) %>% 
  labelled::set_variable_labels(
    imp6_tem = "Tempur Tantrums",
    imp6_res = "Restless",
    imp6_fid = "Fidgeting",
    imp6_dis = "Easily Distracted",
    imp6_thi = "Think Before Act (Reverse Scale)",
    imp6_tas = "Sees Task Through to the End (Reverse Scale)",
    risk6_lie = "Lie or Cheat",
    risk6_ste = "Steal",
    risk6_obe = "Generally Obedient (Reverse Scale)",
    risk6_shl = "Shoplifting",
    risk6_wpp = "Write/Paint in Public Places",
    risk6_smk = "Underage Smoking",
    risk6_drk = "Underage Drinking",
    risk6_rude = "Rude or Noisy in Public",
    risk6_damg = "Damage Something That Doesn't Belong to Them"
  )

# Show descriptive statistics 
self_control6 <- c("imp6_tem", "imp6_res", "imp6_fid", "imp6_dis", "imp6_thi", "imp6_tas","risk6_lie", "risk6_ste", "risk6_obe", "risk6_shl", "risk6_wpp", "risk6_smk", "risk6_drk", "risk6_rude", "risk6_damg")

lapply(self_control6, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  tab1(df5[, var_name], main = var_name)
})
```


Now we do the same for wave 5:

```{r}
df6 <- merged_data %>%
  mutate(
    # Self-Control Variables
    imp5_tem = case_when(
      epsdtt00 %in% c(-1, 4) ~ NA_real_,
      epsdtt00 == 1 ~ 0,
      epsdtt00 == 2 ~ 1,
      epsdtt00 == 3 ~ 2
    ),
    imp5_res = case_when(
      epsdro00 %in% c(-1, 4) ~ NA_real_,
      epsdro00 == 1 ~ 0,
      epsdro00 == 2 ~ 1,
      epsdro00 == 3 ~ 2
    ),
    imp5_fid = case_when(
      epsdfs00 %in% c(-1, 4) ~ NA_real_,
      epsdfs00 == 1 ~ 0,
      epsdfs00 == 2 ~ 1,
      epsdfs00 == 3 ~ 2
    ),
    imp5_dis = case_when(
      epsddc00 %in% c(-1, 4) ~ NA_real_,
      epsddc00 == 1 ~ 0,
      epsddc00 == 2 ~ 1,
      epsddc00 == 3 ~ 2
    ),
    imp5_thi = case_when(
      epsdst00 %in% c(-1, 4) ~ NA_real_,
      epsdst00 == 1 ~ 2,
      epsdst00 == 2 ~ 1,
      epsdst00 == 3 ~ 0
    ),
    imp5_tas = case_when(
      epsdte00 %in% c(-1, 4) ~ NA_real_,
      epsdte00 == 1 ~ 2,
      epsdte00 == 2 ~ 1,
      epsdte00 == 3 ~ 0
    ),
    # Risk Taking Variables
    risk5_lie = case_when(
      epsdoa00 %in% c(-1, 4) ~ NA_real_,
      epsdoa00 == 1 ~ 0,
      epsdoa00 == 2 ~ 1,
      epsdoa00 == 3 ~ 2
    ),
    risk5_ste = case_when(
      epsdcs00 %in% c(-1, 4) ~ NA_real_,
      epsdcs00 == 1 ~ 0,
      epsdcs00 == 2 ~ 1,
      epsdcs00 == 3 ~ 2
    ),
    risk5_obe = case_when(
      epsdor00 %in% c(-1, 4) ~ NA_real_,
      epsdor00 == 1 ~ 2,
      epsdor00 == 2 ~ 1,
      epsdor00 == 3 ~ 0
    ),
    risk5_sch = case_when(
      ecq45x00 == -8 ~ NA_real_,
      ecq45x00 == 2 ~ 0,
      ecq45x00 == 1 ~ 1
    ),
    risk5_shl = case_when(
      ecq26x00 == -8 ~ NA_real_,
      ecq26x00 == 2 ~ 0,
      ecq26x00 == 1 ~ 1
    ),
    risk5_wpp = case_when(
      ecq27x00 == -8 ~ NA_real_,
      ecq27x00 == 2 ~ 0,
      ecq27x00 == 1 ~ 1
    ),
    risk5_smk = case_when(
      ecq59x00 == -8 ~ NA_real_,
      ecq59x00 == 2 ~ 0,
      ecq59x00 == 1 ~ 1
    ),
    risk5_drk = case_when(
      ecq61x00 == -8 ~ NA_real_,
      ecq61x00 == 2 ~ 0,
      ecq61x00 == 1 ~ 1
    ),
    risk5_rude = case_when(
      ecq25x00 == -8 ~ NA_real_,
      ecq25x00 == 2 ~ 0,
      ecq25x00 == 1 ~ 1
    ),
    risk5_damg = case_when(
      ecq28x00 == -8 ~ NA_real_,
      ecq28x00 == 2 ~ 0,
      ecq28x00 == 1 ~ 1
    )
  ) %>%
  dplyr::select(imp5_tem, imp5_res, imp5_fid, imp5_dis, imp5_thi, imp5_tas, risk5_lie, risk5_ste, risk5_obe, risk5_sch, risk5_shl, risk5_wpp, risk5_smk, risk5_drk, risk5_rude, risk5_damg) %>% 
  labelled::set_variable_labels(
    imp5_tem = "Tempur Tantrums",
    imp5_res = "Restless",
    imp5_fid = "Fidgeting",
    imp5_dis = "Easily Distracted",
    imp5_thi = "Think Before Act (Reverse Scale)",
    imp5_tas = "Sees Task Through to the End (Reverse Scale)",
    risk5_lie = "Lie or Cheat",
    risk5_ste = "Steal",
    risk5_obe = "Generally Obedient (Reverse Scale)",
    risk5_sch = "Miss School",
    risk5_shl = "Shoplifting",
    risk5_wpp = "Write/Paint in Public Places",
    risk5_smk = "Underage Smoking",
    risk5_drk = "Underage Drinking",
    risk5_rude = "Rude or Noisy in Public",
    risk5_damg = "Damage Something That Doesn't Belong to Them"
  )


# Show frequency
self_control5 <- c("imp5_tem", "imp5_res", "imp5_fid", "imp5_dis", "imp5_thi", "imp5_tas","risk5_lie", "risk5_ste", "risk5_obe", "risk5_sch", "risk5_shl", "risk5_wpp", "risk5_smk", "risk5_drk", "risk5_rude", "risk5_damg")


lapply(self_control5, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  tab1(df6[, var_name], main = var_name)
})
```


Now, let's merge all the self-contorl variables into one data frame:

```{r}
df_sc <- bind_rows(df4, df5, df6)
# Check dimensions
dim(df_sc)
```


Another concept we are interested in is parenting in childhood. In this study, we are especially interested in parents' behaviors of corrections when their children misbehave. In the following parts, we will look at parenting in wave 5 (age 11) and wave 4 (age 7).

We begin by looking at variables measuring parenting at wave 5 (age 11): 

```{r}
df7 <- merged_data %>%
  mutate(
    par5_plwk = case_when(
      epplwk00 %in% c(-8, -1) ~ NA_real_,
      epplwk00 == 1 ~ 0,
      epplwk00 == 2 ~ 1,
      epplwk00 == 3 ~ 2,
      epplwk00 == 4 ~ 3,
      epplwk00 == 5 ~ 4
    ),
    par5_dibn = case_when(
      epdibn00 %in% c(-1, 6) ~ NA_real_,
      epdibn00 == 1 ~ 0,
      epdibn00 == 2 ~ 1,
      epdibn00 == 3 ~ 2,
      epdibn00 == 4 ~ 3,
      epdibn00 == 5 ~ 4
    ),
    par5_ditr = case_when(
      epditr00 %in% c(-1, 6) ~ NA_real_,
      epditr00 == 1 ~ 0,
      epditr00 == 2 ~ 1,
      epditr00 == 3 ~ 2,
      epditr00 == 4 ~ 3,
      epditr00 == 5 ~ 4
    ),
    par5_dire = case_when(
      epdire00 %in% c(-1, 6) ~ NA_real_,
      epdire00 == 1 ~ 0,
      epdire00 == 2 ~ 1,
      epdire00 == 3 ~ 2,
      epdire00 == 4 ~ 3,
      epdire00 == 5 ~ 4
    )
  ) %>% 
  dplyr::select(par5_plwk, par5_dibn, par5_ditr, par5_dire) %>% 
  labelled::set_variable_labels(
  par5_plwk = "Unsupervised hours with friends",
  par5_dibn = "Send to bedrooms when misbehaves",
  par5_ditr = "Take away treats when misbehaves",
  par5_dire = "Try to reason with her when misbehaves"
)

# Show frequency
parenting5 <- c("par5_plwk", "par5_dibn", "par5_ditr", "par5_dire")

lapply(parenting5, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  print(summary(df7[[var_name]]))
  tab1(df7[, var_name], main = var_name)
})

```


Now, we do the same for wave 4 (age 7)

```{r}
df8 <- merged_data %>%
  mutate(
    # Send to bedrooms when misbehaves
    par4_dibn = case_when(
      dmdibna0 %in% c(-9, -8, -1, 6) ~ NA_real_,
      dmdibna0 == 1 ~ 0,
      dmdibna0 == 2 ~ 1,
      dmdibna0 == 3 ~ 2,
      dmdibna0 == 4 ~ 3,
      dmdibna0 == 5 ~ 4
    ),
    # Take away treats when misbehaves
    par4_ditr = case_when(
      dmditra0 %in% c(-9, -8, -1, 6) ~ NA_real_,
      dmditra0 == 1 ~ 0,
      dmditra0 == 2 ~ 1,
      dmditra0 == 3 ~ 2,
      dmditra0 == 4 ~ 3,
      dmditra0 == 5 ~ 4
    ),
    # Try to reason with her when misbehaves
    par4_dire = case_when(
      dmdirea0 %in% c(-9, -8, -1, 6) ~ NA_real_,
      dmdirea0 == 1 ~ 0,
      dmdirea0 == 2 ~ 1,
      dmdirea0 == 3 ~ 2,
      dmdirea0 == 4 ~ 3,
      dmdirea0 == 5 ~ 4
    )
  ) %>% dplyr::select(par4_dibn, par4_ditr, par4_dire) %>%
  labelled::set_variable_labels(
  par4_dibn = "Send to bedrooms when misbehaves",
  par4_ditr = "Take away treats when misbehaves",
  par4_dire = "Try to reason with her when misbehaves"
)

# Display summary for each variable
parenting4 <- c("par4_dibn", "par4_ditr", "par4_dire")

lapply(parenting4, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  print(summary(df8[[var_name]]))
  tab1(df8[, var_name], main = var_name)
})
```


Now, we can combine all parenting variables into one data frame:

```{r}
df_pa <- bind_rows(df7, df8)
# Check dimensions
dim(df_pa)
```


Finally, let's look at and recode the potential covariates:

```{r}
df_co <- merged_data %>%
  mutate(
    pared = case_when(
      fdacaq00 == 96 ~ 0,
      fdacaq00 == 95 | fdacaq00 == -1 ~ NA_real_,
      TRUE ~ fdacaq00
    ),
    
    race = case_when(
      fd06e00 %in% c(-9, -1) ~ NA_real_,
      fd06e00 == 1 ~ 1,
      fd06e00 == 2 ~ 4,
      fd06e00 %in% c(3, 4) ~ 2,
      fd06e00 == 5 ~ 3,
      fd06e00 == 6 ~ 4
    ),
    
    gender6 = case_when(
      fccsex00 == 2 ~ 0,
      TRUE ~ 1
    ),
    
    married = case_when(
      amfcin00 %in% c(-9, -8, -1) ~ NA_real_,
      amfcin00 %in% c(1, 4, 5, 6) ~ 0,
      amfcin00 %in% c(2, 3) ~ 1
    )
  ) %>%
  labelled::set_variable_labels(
    pared = "Highest Parent's Education",
    race = "Racial Categories",
    gender6 = "Gender of CM",
    married = "Marital Status at 9 month-old"
  ) %>%
  mutate(
    race = factor(race, levels = c(1, 2, 3, 4), labels = c("White British", "Asian British", "Black British", "Other British")),
    gender = factor(gender6, levels = c(0, 1), labels = c("Female", "Male"))
  ) %>% 
  dplyr::select(pared, race, gender, married)

# Show frequency
covar <- c("pared", "race", "gender", "married")

lapply(covar, function(var_name) {
  cat("\nVariable:", var_name, "\n")
  print(summary(df_co[[var_name]]))
  tab1(df_co[, var_name], main = var_name)
  
})
```


Finally, let's combine all the relevant columns into one data frame:

```{r}
df_re <- bind_rows(df_aggr, df_sc, df_co, df_pa)
# Check dimensions
dim(df_re)
# Write it to r data file (uncomment when needed)
## saveRDS(df_re, file = "/home/freeyang/research_data/mcs/merging/relevant_columns.rds")
```


## Potential Issues of Data

From the above analysis we identified the following potential issues with our dataset:


1. **Lots of missing cases**

This is not uncommon when working with panel data. Most variables in wave 7 (age 17) have about 40%-50% missing. The missing cases are very likely to be related to attrition. However, we suggest careful missing value analysis before proceeding. 

2. **Sample Unbalance**

Another critical issue we identified from the above sections is the data unbalance issue. Most variables we plan to use are highly unbalanced. Specifically, most variables measuring adolescent aggressive behaviors are extremely unbalance. Some of the variables measuring adolescent aggressive behaviors only have about 2% positive cases. Unbalanced endogenous variables require special treatment in conventional SEM. How extremely unbalanced endogenous variables affect tree-based SEM also requires further attention. 




















