---
title: "Data Preperation"
author: "Siyang Ni"
date: "2024-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Description

The dataset is from a longgitudinal social survey in the UK called the "Millennium Cohort Study" (MCS). The MCS follows the lives of around 19,000 young people born in England, Scotland, Wales, and Northern Ireland in 2000-02. 

MCS provides multiple measures of the cohort members’ physical, socio-emotional, cognitive and behavioral development over time, as well as detailed information on their daily life, behavior and experiences. Alongside this, rich information on economic circumstances, parenting, relationships and family life is available from both resident parents.

Currently, the MCS have seven waves available to researchers. The first wave of survey was taken at the year 2001, when respondents were nine-month old. The seventh wave survey was taken at 2018, when the respondents were seventeen-year-old.   


## Tasks

We will conduct the following task:

1. Merge the seven waves of data.
2. Provide a general description of the contents of your sample.
3. Identify issues that may need to be addressed in later analyses.
4. Exploratory data visualization 
5. Some very basic exploratory data analysis

Computing Environment: 

-Program: R 4.3.2
-IDE: R Studio Server 2023.12.1+402 "Ocean Storm" Release 
-OS: Ubuntu 22.04.4 LTS through WSL2 (x86_64-pc-linux-gnu)


## Loading and Merging 

```{r}
setwd("/home/freeyang/research/hdfs597")

# install.packages('pacman')
# Pacman is an efficient R package manager, works like APT in Debian
library(pacman) 

# For merging we need haven and dplyrgit config --global user.email "youremail@domain.com"
pacman::p_load(haven, dplyr)
```

```{r}
# Load all waves

## Base directory prefix
base_dir <- "/home/freeyang/research_data/mcs"

# Reading each wave, all column names to lowercase, filter out twins and triplets, keep only main caregivers. 

# Wave 1 Parent Interview
wave1_parent_interview <- read_dta(file.path(base_dir, "MCS 1/stata11/mcs1_parent_interview.dta")) %>%
  rename_with(tolower)

# Wave 2 Child Assessment
wave2_child_assessment <- read_dta(file.path(base_dir, "MCS 2/stata11_se/mcs2_child_assessment_data.dta")) %>%
  rename_with(tolower) %>%
  filter(bhcnum00 == 1)

# Wave 2 Parent Interview
wave2_parent_interview <- read_dta(file.path(base_dir, "MCS 2/stata11_se/mcs2_parent_interview.dta")) %>%
  rename_with(tolower)

# Wave 3 Child Assessment
wave3_child_assessment <- read_dta(file.path(base_dir, "MCS 3/stata11_se/mcs3_child_assessment_data.dta")) %>%
  rename_with(tolower) %>%
  filter(chcnum00 == 1)

# Wave 3 Parent Interview
wave3_parent_interview <- read_dta(file.path(base_dir, "MCS 3/stata11_se/mcs3_parent_interview.dta")) %>%
  rename_with(tolower)

# Wave 4 Parent Interview
wave4_parent_interview <- read_dta(file.path(base_dir, "MCS 4/stata11_se/mcs4_parent_interview.dta")) %>%
  rename_with(tolower)

# Wave 4 Child Assessment
wave4_child_assessment <- read_dta(file.path(base_dir, "MCS 4/stata11_se/mcs4_cm_assessment.dta")) %>%
  rename_with(tolower) %>%
  filter(dccnum00 == 1)

# Wave 5 CM Assessment
wave5_cm_assessment <- read_dta(file.path(base_dir, "MCS 5/stata11/mcs5_cm_cognitive_assessment.dta")) %>%
  rename_with(tolower) %>%
  filter(ecnum00 == 1)

# Wave 5 CM Interview
wave5_cm_interview <- read_dta(file.path(base_dir, "MCS 5/stata11/mcs5_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(ecnum00 == 1)

# Wave 5 Parent CM Interview
wave5_parent_cm_interview <- read_dta(file.path(base_dir, "MCS 5/stata11/mcs5_parent_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(ecnum00 == 1, epnum00 == 1)

# Wave 5 Child Derived
wave5_child_derived <- read_dta(file.path(base_dir, "MCS 5/stata11/mcs5_cm_derived.dta")) %>%
  rename_with(tolower) %>%
  filter(ecnum00 == 1)

# Wave 6 CM Interview
wave6_cm_interview <- read_dta(file.path(base_dir, "MCS 6/stata11/mcs6_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(fcnum00 == 1)

# Wave 6 Parent CM Interview (Main)
wave6_parent_cm_interview <- read_dta(file.path(base_dir, "MCS 6/stata11/mcs6_parent_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(fpnum00 == 1, fcnum00 == 1)

# Wave 6 Parent Derived
wave6_parent_derived <- read_dta(file.path(base_dir, "MCS 6/stata11/mcs6_parent_derived.dta")) %>%
  rename_with(tolower) %>%
  filter(fpnum00 == 1)

# Wave 6 Child Derived
wave6_child_derived <- read_dta(file.path(base_dir, "MCS 6/stata11/mcs6_cm_derived.dta")) %>%
  rename_with(tolower) %>%
  filter(fcnum00 == 1)

# Wave 7 CM Interview
wave7_cm_interview <- read_dta(file.path(base_dir, "MCS 7/stata13/mcs7_cm_interview_onlyCM1.dta")) %>%
  rename_with(tolower)

# Wave 7 Parent CM Interview
wave7_parent_cm_interview <- read_dta(file.path(base_dir, "MCS 7/stata13/mcs7_parent_cm_interview.dta")) %>%
  rename_with(tolower) %>%
  filter(gpnum00 == 1, gcnum00 == 1)

# Wave 7 CM Derived
wave7_cm_derived <- read_dta(file.path(base_dir, "MCS 7/stata13/mcs7_cm_derived_onlyCM1.dta")) %>%
  rename_with(tolower)
```

```{r}
# Merge all waves

## purrr package for merging
pacman::p_load(purrr)

## Store all waves in a list
data_frames_list <- list(
  wave1_parent_interview,
  wave2_child_assessment,
  wave2_parent_interview,
  wave3_child_assessment,
  wave3_parent_interview,
  wave4_parent_interview,
  wave4_child_assessment,
  wave5_cm_assessment,
  wave5_cm_interview,
  wave5_parent_cm_interview,
  wave5_child_derived,
  wave6_cm_interview,
  wave6_parent_cm_interview,
  wave6_parent_derived,
  wave6_child_derived,
  wave7_cm_interview,
  wave7_parent_cm_interview,
  wave7_cm_derived
)

## Use reduce from purrr to merge all waves 1 by 1 
merged_data <- purrr::reduce(data_frames_list, full_join, by = "mcsid")

## Save the merged dataset as an RDS file
### (Uncomment when needed)
#(merged_data, file = "/home/freeyang/research_data/mcs/merging/merged_dataset.rds")
```


## Data Exploration

A quick look shows we have a very large dataset with 19,243 rows and 18,816 columns, and also many missing cases. This is expected for a panel data that span from when the respondents were nine-month-old to seventeen-year-old.

```{r}
## Load the merged dateset\
merged_data <- readRDS(file = "/home/freeyang/research_data/mcs/merging/merged_dataset.rds")

## Summarize the data
dim(merged_data)
head(merged_data[1:5, 1:10])

## Missing Values
total_missing_values <- sum(is.na(merged_data))
cat("Total Missing Values:", total_missing_values, "\n")
```


Our main target variable is adolescent aggressive behaviors. We look at these variables in each waves:

```{r}
# Show frequency distribution for variablrs measuring aggression at age 17
aggg_7 <- c("gpsdfb00", "gchitt00", "gcwepn00", "gcsdql00")

freq_tables <- lapply(merged_data[aggg_7], table)

lapply(names(freq_tables), function(x) {
  cat("\nFrequency table for", x, ":\n")
  print(freq_tables[[x]])
})
```
































```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
